The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-12 rational.tex
whatever

Applying a function consists of making it flow backwards to satisfy the demand 
of the function:
\begin{multicols}{2}
\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/apply1.pdf}
}
\end{center}
\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/apply2.pdf}
}
\end{center}
\end{multicols}

@@

Most of the rules look obviously correct but there is a subtle
point. Consider the case for {{c1 (+) c2}}. Previously the shape of the value
determined which of the combinators {{c1}} or {{c2}} to use. Now the incoming
value could be a fresh logical variable, and indeed we have two rules with
the same left hand side, and only depending on the success or failure of some
further unification, can we decide which of them applies. The situation is
common in logic programming in the sense that it is considered a
non-deterministic process that keeps searching for the right substitution if
any. Typical implementations of logic programming languages further provide
top-level mechanisms to manipulate this non-deterministic search, for example
by returning one answer and giving the user to ask for more answers. We
abstract from this top-level semantics and view the rules above as being
applied non-deterministically.

@@

Given that our language is reversible (Proposition~\ref{prop:logrev}), a
backward evaluator is relatively straightforward to implement: it simply
corresponds to. Indeed using the backward evaluator to calculate {{c v}} is
equivalent {{c^{dagger} v}} in the forward evaluator.

@@

    \item The rules for {{eta+}} are added to backward evaluator. A
      program executing backward starts executing forward after the
      execution of the {{eta+}}. Dual to the previous case, there is
      no rule for {{eps+}} in the backward evaluator since the output
      type of {{eps+}} is {{0}}.
%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% <eta+, v, C, s>_{bck} |--> <eta+, left (-v'), C, s[v <><> right v']>_{fwd}
% <eta+, v, C, s>_{bck} |--> <eta+, right v', C, s[v <><> left (-v')]>_{fwd}

@@

With a little work, we can modify this circuit to only annihilate values that
fail to satisfy the constraints reprsented by a SAT-instance {{f}}. In more
detail, an instance of SAT is a function/circuit {{f}} that given some
boolean inputs returns {{true}} or {{false}} which we interpret as whether
the inputs satisfy the constraints imposed by the structure of {{f}}. Because
we are in a reversible world, our instance of SAT must be expressed as an
isomorphism: this is easily achieved as shown in Sec.~\ref{sub:f}
below. Assuming that {{f}} is expressed as an isomorphism, we have enough
information to reconstruct the input from the output. This can be done by
using the adjoint of {{f}}. At this point we have, the top half of the
construction below:

>>>>

Applying a function to an argument consists of making the argument flow
backwards to satisfy the demand of the function:
\begin{multicols}{2}
\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/apply1.pdf}
}
\end{center}
\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/apply2.pdf}
}
\end{center}
\end{multicols}

@@

Most of the rules look obviously correct but there is a subtle
point. Consider the case for {{c1 (+) c2}}. Previously the shape of the value
determined which of the combinators {{c1}} or {{c2}} to use. Now the incoming
value could be a fresh logical variable, and indeed we have two rules with
the same left hand side, and only depending on the success or failure of some
further unification, can we decide which of them applies. The situation is
common in logic programming in the sense that it is considered a
non-deterministic process that keeps searching for the right substitution if
any. Typical implementations of logic programming languages further provide
top-level mechanisms to manipulate this non-deterministic search, for example
by returning one answer and giving the user the option to ask for more
answers. We abstract from this top-level semantics and view the rules above
as being applied non-deterministically.

@@

Given that our language is reversible (Prop.~\ref{prop:logrev}), a backward
evaluator is relatively straightforward to implement: using the backward
evaluator to calculate {{c v}} is equivalent {{c^{dagger} v}} in the forward
evaluator.

@@

    \item The rules for {{eta+}} are added to the backward evaluator. A
      program executing backward starts executing forwards after the
      execution of the {{eta+}}. Dual to the previous case, there is
      no rule for {{eps+}} in the backward evaluator since the output
      type of {{eps+}} is {{0}}.
%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% <eta+, v, C, s>_{bck} |--> <eta+, left (-v'), C, s[v <><> right v']>_{fwd}
% <eta+, v, C, s>_{bck} |--> <eta+, right v', C, s[v <><> left (-v')]>_{fwd}

@@

With a little work, we can modify this circuit to only annihilate values that
fail to satisfy the constraints reprsented by a SAT-instance {{f}}. In more
detail, an instance of SAT is a function/circuit~{{f}} that given some
boolean inputs returns {{true}} or {{false}} which we interpret as whether
the inputs satisfy the constraints imposed by the structure of {{f}}. Because
we are in a reversible world, our instance of SAT must be expressed as an
isomorphism: this is easily achieved as shown in Sec.~\ref{sub:f}
below. Assuming that {{f}} is expressed as an isomorphism, we have enough
information to reconstruct the input from the output. This can be done by
using the adjoint of {{f}}. At this point we have, the top half of the
construction below:
